MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-08-02 main/MSstatsPTM_Main.tex
eval comments

@@

We compared the performance of  $MSstatsPTM$ to that of \todo{Need more details on the 'adjusted' vs 'unadjusted' versions ANOVA and Limma. Give each version a name that you'll later use in the figures, and describe what type of summaries each version takes.} ANOVA and of Limma. Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons.  $MSstatsPTM$ \todo{Also describe the 'adjusted' version (MSstatsPTM) and the 'unadjusted' version (I guess it is MSstats/MSstatsTMT)} was used without imputing missing values, and without Empirical Bayes moderation. All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

@@

\figref{sim1_fdr}(a) summarizes the FDR \todo{I am not sure we should call this FDR. By definition FDR is a population-level quantity (it is an expected value of the proportion of false positives. Here it is an observed value). Maybe replace with 1-PPV?} reported on Computer Simulation 1 dataset by the unadjusted and adjusted versions of methods \todo{XXX}. All the analyses were performed to control the FDR at at most 5\%.  The  simulation mimics a ``clean" label-free experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers. Yet, even under these favorable circumstances, the models that did not adjust for confounding from changes in overall protein abundance produced an excessive number of false positives. The versions of the models that accounted for the confounding produced error rates that were much better calibrated at the desired level. Since the simulation mimicked the assumptions of \todo{model XXX}, the error rates of this model were calibrated best.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In noisy simulations, MSstatsPTM improved the estimation of fold change over the existing methods \todo{Hmm... Fig 4b is not about the fold change... Rephrase?}}

\todo{I suggest this order for this section. (Fig 4b has multiple panels, but the current narrative doesn't make it quite clear what is the difference between the panels) Fig 4b summarizes the results on Sim 2 dataset. The dataset is more realistic, and includes XXX. For low noise, methods compared in terms of sample size and number of conditions XXX. As the noise increased, XXX. The difference is due primarily to the robust nature of summarization in MSstats. Supplementary Fig XXX details the estimation of fold change}

In simulations with missing values and few features, $MSstatsPTM$ outperformed ANOVA and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy  well, even when the number of replicates were low. $MSstatsPTM$ can handle low feature counts better than the existing methods due to using Tukey's Median Polish for summarization. The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the existing method's distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and ANOVA's IQR. While Limma and ANOVA overall estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate. The proposed method's estimation was more consistent, with a smaller error. As discussed in the previous section, the summarization methods of the proposed methods excel when the number of observations are low. This was the case here, with each PTM only averaging $1.37$ features. The existing approaches used a non-robust summarization method, that was very sensitive to few observations, causing their fold change estimation to be more inconsistent, with a larger error compared to the proposed method.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.7\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:sim1_fdr}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\label{fig:sim2_acc}
\end{subfigure}
\caption{\todo{Could you clarify what 'with' and 'without' adjustment means for each model? E.g., what is 'proposed with no adjustment'? We did not propose anything without adjustment - did you mean base MSstats? Also, maybe clarify what you mean by limma with and without adjustment, since we technically didn't introduce limma without the adjustment. Maybe explain the inputs to each version of the model?} \todo{In (a), could you add a horizontal line at 5\%?} Computer simulations. a) Dataset 1, clean simulation, analyzed to control FDR at at most 5\%. The methods not accounting for the confounding between changes in PTM and overall changes in protein abundance produced exceedingly high numbers of false positives. In contrast, the methods accounting for the confounding correctly calibrated the proportion of false positive differentially abundant PTM.  b) In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent. Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_boxplot}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free. a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models. After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and ANOVA models was wider than the proposed method. b) \todo{The title of the figure reads as if it is accuracy-adjusted. Maybe rephrase?} Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation. Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
\label{fig:spikein_volcano}
\end{figure}

>>>>

We compared the performance of $MSstatsPTM$ to that of$Limma$ and $ANOVA$. Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons. All methods were compared both before and after adjusting for changes in the unmodified protein. $MSstatsPTM$ without adjustment is the same as running $MSstats$ on the peptide level. $ANOVA$ without adjustment is shown in \eqref{ttest_model}, and will be referred to as basic $ANOVA$. Finally, $Limma$ without adjustment uses the same model formula in \eqref{ttest_model}, while including a moderated variance estimation, and is referred to as basic $Limma$. $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation. All the evaluations were done at the FDR-adjusted p-value cutoff of $q=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

@@

@@

\figref{sim1_fdr}(a) summarizes the FDR \todo{I am not sure we should call this FDR. By definition FDR is a population-level quantity (it is an expected value of the proportion of false positives. Here it is an observed value). Maybe replace with 1-PPV?} reported on Computer Simulation 1 dataset by $MSstatsPTM$, $ANOVA$, $Limma$ and the basic $MSstats$, $ANOVA$ and $Limma$ methods. All the analyses were performed to control the FDR at at most 5\%.  The  simulation mimics a ``clean" label-free experiment, not compromised by issues such as deviations from model assumptions, missing values and outliers. Yet, even under these favorable circumstances, the models that did not adjust for confounding from changes in overall protein abundance produced an excessive number of false positives. The versions of the models that accounted for the confounding produced error rates that were much better calibrated at the desired level. Since the simulation mimicked the assumptions of \todo{model XXX}, the error rates of this model were calibrated best.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In noisy simulations, MSstatsPTM more accurately detected differentially modified modifications over the existing methods}

Figure \figref{sim2_acc} summarizes the results of the second computer simulation. The simulation is more realistic, and includes low modified feature counts and missing values. Differing numbers of conditions and replicates were compared in the presence of low and high noise. For low noise, $MSstatsPTM$ outperformed the existing methods across all conditions and number of replicates, with near 100\% accuracy when the replicates were high. As the noise increased the accuracy of all the methods decreased, however $MStatsPTM$ still outperformed the existing methods. The difference is primarily due to the robust nature of summarization in $MSstatsPTM$ through Tukey's Median Polish. The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however $MSstatsPTM$'s distribution of estimated fold changes was visibly tighter than the existing methods (\figref{spikein_boxplot}). Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and ANOVA's IQR. While Limma and ANOVA generally estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate. The proposed method's estimation was more consistent, with a smaller error. As discussed in the previous section, the summarization methods of the proposed methods excel when the number of observations are low. This was the case here, with each PTM only averaging $1.37$ features. The existing approaches used a non-robust summarization method, that was very sensitive to few observations, causing their fold change estimation to be more inconsistent, with a larger error compared to the proposed method.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.7\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:sim1_fdr}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\label{fig:sim2_acc}
\end{subfigure}
\caption{\todo{Could you clarify what 'with' and 'without' adjustment means for each model? E.g., what is 'proposed with no adjustment'? We did not propose anything without adjustment - did you mean base MSstats? Also, maybe clarify what you mean by limma with and without adjustment, since we technically didn't introduce limma without the adjustment. Maybe explain the inputs to each version of the model?} \todo{In (a), could you add a horizontal line at 5\%?} Computer simulations. a) Dataset 1, clean simulation, analyzed to control FDR at at most 5\%. The methods not accounting for the confounding between changes in PTM and overall changes in protein abundance produced exceedingly high numbers of false positives. In contrast, the methods accounting for the confounding correctly calibrated the proportion of false positive differentially abundant PTM.  b)  \todo{The title of the figure reads as if it is accuracy-adjusted. Maybe rephrase?} In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent. Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_boxplot}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free. a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models. After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and ANOVA models was wider than the proposed method. b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation. Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
\label{fig:spikein_volcano}
\end{figure}
